Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-04-29 manuscript/manuscript.tex
added P-P plot

\author{Xavier Siemens}
\affiliation{Center for Gravitation, Cosmology and Astrophysics, University of Wisconsin--Milwaukee, PO Box 413, Milwaukee WI, 53201, USA}

@@

The fixed-noise and noise-marginalized analyses give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
%Figure~\ref{fig:os_datasetstats2} shows the mean optimal statistic and mean SNR 
%for 100 realizations of a GWB with $\Agw = 10^{-15}$. 
%The fixed-noise analysis using the individual noise values 
%gives a mean value of $\hat{A}^2 = 2.9\times10^{-31}$ and a mean SNR of 1.1. 
%The fixed-noise analysis using the common noise values 
%gives a mean value of $\hat{A}^2 = 8.4\times10^{-31}$ and a mean SNR of 2.1, 
%which the noise-marginalized analysis gives a mean value of 
%$\hat{A}^2 = 1.1\times10^{-30}$ and a mean SNR of 2.4.
%\begin{figure}[t]
%	\includegraphics[width=0.95\columnwidth]{plots/optstat_A1e-15.pdf}
%	\caption{Optimal statistic and SNR for 100 simulated datasets 
%			containing an injected GWB with $\hat{A}^2 = 10^{-30}$ (dashed vertical line). 
%			We show the values found by fixing the red noise 
%			based on the results from individual and common noise analyses 
%			(blue and orange lines, respectively) 
%			to the mean optimal statistic and mean SNR found by marginalizing over 
%			the red noise parameters using the results of a common noise analysis (green lines). 
%			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
%			that are significantly lower than the injected value, 
%			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
%			that are closer to the injected value.}
%	\label{fig:os_datasetstats2}
%\end{figure}
We also compare these three methods

Figure~\ref{fig:os_compare} shows the distribution of $(\hat{A}^2-\Agw^2)/\sigma$
\begin{figure*}[t]
	\includegraphics[width=0.95\textwidth]{plots/pp_plot.pdf}
	\caption{P--P plot showing the cumulative fraction of simulations for which $\Agw^2$ lies within 
			a given confidence interval of the measured $\hat{A}^2$. 
			The probability distribution of $\hat{A}^2$ is assumed to be a Gaussian 
			with variance $\sigma^2_{\hat{A}^2}$. 
			The fixed-noise optimal statistic using the individual and common noise results 
			both give biased values of $\hat{A}^2$, particularly for small values of \Agw, 
			while the noise-marginalized optimal statistic gives more accurate values of $\hat{A}^2$ 
			over a large range of injected values of \Agw.}
	\label{fig:os_compare}
\end{figure*}

@@

\acknowledgments
We thank Joe Romano and Jeff Hazboun for useful discussions. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA.

>>>>

%\author{Xavier Siemens}
%\affiliation{Center for Gravitation, Cosmology and Astrophysics, University of Wisconsin--Milwaukee, PO Box 413, Milwaukee WI, 53201, USA}

@@

The fixed-noise and noise-marginalized analyses give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$, 
assuming a Gaussian distribution with mean equal to the measured $\hat{A}^2$ 
and variance $\sigma^2_{\hat{A}^2}$, 
determined from the distribution of $\hat{A}^2$ for our simulations. 
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope 1 (the dotted, diagonal line).

We compare the three methods for computing the optimal statistic for 
simulations with $\Agw=5\times10^{-15}$, $\Agw=10^{-15}$, and $\Agw=5\times10^{-16}$. 
The fixed-noise optimal statistic using the individual noise results systematically underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, left panel). 
The fixed-noise optimal statistic using the common noise results 
estimates $\hat{A}^2$ well for large values of $\Agw$, but for small values it also underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, middle panel). 
The noise-marginalized optimal statistic provides the most accurate estimate of $\hat{A}^2$ 
over the range of \Agw\ considered here (Fig.~\ref{fig:os_compare}, right panel).
\begin{figure*}[t]
	\includegraphics[width=0.95\textwidth]{plots/pp_plot.pdf}
	\caption{P--P plot showing the cumulative fraction of simulations for which $\Agw^2$ lies within 
			a given confidence interval of the measured $\hat{A}^2$. 
			The probability distribution of $\hat{A}^2$ is assumed to be a Gaussian 
			with variance $\sigma^2_{\hat{A}^2}$. 
			The fixed-noise optimal statistic using the individual and common noise results 
			both give biased values of $\hat{A}^2$, particularly for small values of \Agw, 
			while the noise-marginalized optimal statistic gives more accurate values of $\hat{A}^2$ 
			over a large range of injected values of \Agw.}
	\label{fig:os_compare}
\end{figure*}

@@

\acknowledgments
We thank Joe Romano, Xavier Siemens, Jolien Creighton, and Jeff Hazboun for useful discussions. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA.
